---
title: "Healthcare Cost and Utilization Project"
author: "Akashdeep Jaswal"
date: "10/31/2016"
output: pdf_document
---

### About Healthcare Cost and Utilization Project (HCUP):

HCUP is the Nation’s most comprehensive source of hospital data, including information on in-patient care, ambulatory care, and emergency department visits. HCUP enables researchers, insurers, policymakers and others to study health care delivery and patient outcomes over time, and at the national, regional, State, and community levels. For details about this initiative [see](http://www.hcup-us.ahrq.gov/)


```{r, message=FALSE, warning=FALSE}
# Importing Packages
library(tidyverse)
library(plm)
library(ggplot2)
library(lattice)

```

### 1. Importing Data

#### 1.1 Reading Inpatient Hospitalization Data

Variable encoding scheme and definitions for HCUP Data: [here](https://www.hcup-us.ahrq.gov/db/state/siddist/sid_multivar.jsp)

```{r}
setwd("~/Desktop/AkashPersonal/hcup-population")

# Reading the Hospitalization Data
hosp.data <- NULL
for (year in seq(2004,2013,by=1)) {
  
  # path name on how to read the file  
  file <- paste('data/HospData/WA_SIDC_',year,'_CORE','.rdat',sep='')
  print (c("Reading data from", file))
  
  # Loading RDat files and Appending to DF 
  load(file)
  cols <- c("ayear","amonth","pstco2","zip","pl_ur_cat4","age","female","dx1","dxccs1")
  temp <- data.frame(temp)[cols]
  hosp.data <- rbind(hosp.data,temp)
  
  # Tabular Dataframe
  hosp.data <- tbl_df(hosp.data)
}

```

#### 1.2 Importing Vaccination Data and Census Demographics

```{r}

vaccine.data <- read.csv('~/Desktop/AkashPersonal/hcup-population/data/Vaccine_clean_data.csv')
census.data <- read.csv('~/Desktop/AkashPersonal/hcup-population/data/Final_Population_Interpolated.csv')

```

### 2. Data Wrangling

#### 2.1 Data Type Casting

```{r, warning=FALSE}
# Recasting Variables appropriately - Hospitalization
hosp.data$zip <- as.integer(hosp.data$zip)
hosp.data$dx1 <- as.integer(hosp.data$dx1)
hosp.data$pl_ur_cat4 <- as.integer(hosp.data$pl_ur_cat4)
hosp.data <- hosp.data %>% 
  mutate(urban_rural=ifelse(pl_ur_cat4==1, "Large Metropolitan", 
                            ifelse(pl_ur_cat4==2, "Small Metropolitan", 
                            ifelse(pl_ur_cat4==3, "Micropolitan",
                                   "Not Metro/Micro"))))
hosp.data$urban_rural <- as.factor(hosp.data$urban_rural)

# Demographic data from US Census (Monthly Interpolated by WA County)
census.data$year <- as.integer(census.data$year)

```

#### 2.2 Filtering, Aggregating and Mapping Data

```{r}

# Filtering, Aggregating and Summarizing Hosp Data
hosp.filtered <- hosp.data %>% 
  filter(age >= 0 & age <= 4 & ayear >= 2004) %>%
  mutate(AgeGroup=1) %>%
  group_by(ayear, amonth, zip, pstco2, dx1, dxccs1, AgeGroup, urban_rural) %>%
  summarise(hospitalizations=n())

# Zip Code - County FIPS Mapping
zip.fips.mapping <- hosp.data %>% 
  select(zip, pstco2) %>%
  group_by(zip, pstco2) %>%
  summarise(n())

# Mapping the FIPS County codes to Zip Codes in Vaccine Data
vaccine.data <- merge(vaccine.data, zip.fips.mapping[c('zip','pstco2')], 
                      by.x=c("ADDRESS_ZIP"), by.y=c("zip"))
vaccine.data <- vaccine.data[ , !duplicated(colnames(vaccine.data))]

# Cleaning Census Data
census.clean <- census.data %>%
  select(year, month, County, FIPS, AgeGroup, Population) %>%
  group_by(year, month, County, FIPS, AgeGroup) %>%
  summarise(Population=sum(Population))

```

### 3. Building Fixed Effect Models

#### 3.1 All Cause Hospitalization - Vaccination Fixed Effects

```{r}

# All Cause Hospitalization
all.cause.hosp <- hosp.filtered %>%
  select(ayear, amonth, AgeGroup, hospitalizations, pstco2, urban_rural) %>%
  group_by(ayear, amonth, pstco2, AgeGroup,urban_rural) %>%
  summarise(patients=sum(hospitalizations))

all.cause.hosp <- merge(all.cause.hosp, census.clean, by.x=c("ayear","amonth","pstco2","AgeGroup"),
                       by.y=c("year","month","FIPS","AgeGroup"))
all.cause.hosp <- mutate(all.cause.hosp, hosp_norm = patients/Population)

names(all.cause.hosp)[1] <- 'year'
names(all.cause.hosp)[2] <- 'month'
names(all.cause.hosp)[3] <- 'FIPS'

vaccine.filtered <- vaccine.data %>%
  select(year, month, FIPS=pstco2, X4313314.SERIES, TOTAL.POPULATION) %>%
  group_by(year, month, FIPS) %>%
  summarise(X4313314_Norm=sum(X4313314.SERIES)/sum(TOTAL.POPULATION),
            X4313314=sum(X4313314.SERIES), TOTAL_POPULATION=sum(TOTAL.POPULATION))

all.cause.merge <- merge(all.cause.hosp, vaccine.filtered, by=c("year","month","FIPS"))
all.cause.merge$t_index <- interaction(all.cause.merge$year, all.cause.merge$month,sep=".")
summary(all.cause.merge)
```

```{r, message=FALSE, warning=FALSE}
# Plotting the data
ggplot(all.cause.merge, aes(X4313314_Norm,hosp_norm)) + geom_point(alpha=0.2) +
  geom_smooth() + theme_minimal(base_size = 8)

xyplot(hosp_norm ~ X4313314_Norm  | factor(County), data=all.cause.merge, 
       pch=19, type=c("p","r"), alpha = 0.2, 
       main=list(label="All Cause Hospitalization Trends by County Category",cex=0.8),
       sub=list(label="11",cex=0.1),
       par.settings=simpleTheme(col="#59ABE3", col.line="red"),
       par.strip.text=list(cex=0.5))

```

```{r, message=FALSE, warning=FALSE}
# Fixed Effects All Cause Hospitalization
fe.all.cause <- plm(hosp_norm ~ X4313314_Norm, data = all.cause.merge, 
                    index = c("FIPS", "t_index"), model = "within", effect = "individual")

summary(fe.all.cause)
summary(fixef(fe.all.cause))
fixef.all.cause <- matrix(fixef(fe.all.cause))
fixef.df <- as.data.frame(cbind(sort(unique(all.cause.merge$FIPS)), fixef.all.cause))
colnames(fixef.df) <- c("FIPS","coeff")
fixef.df$FIPS <- as.character(fixef.df$FIPS)
fixef.df <- mutate(fixef.df, id=substr(FIPS, 3,5))
fixef.df <- fixef.df[c("coeff",'id')]

```


```{r message=FALSE, message=FALSE, warning=FALSE}
# Setting up Mapping Libraries
library(GGally)
library(rgdal)
library(ggmap)
library(rgeos)
library(broom)
library(Cairo)
library(maptools)
library(maps)
library(choroplethr)

```

```{r, message=FALSE, warning=FALSE}

# Loading Shapefiles and boundaries
map.shapefile <- readOGR(dsn = "/Users/ajaswal/Desktop/hcup-population/Shapefiles/tl_2013_53_cousub", 
                         layer = "tl_2013_53_cousub")
map.df <- fortify(map.shapefile, region="COUNTYFP") 
```

```{r, message=FALSE, warning=FALSE}
map.df.merge <- merge(map.df,fixef.df, by="id") 
ggplot() +
  geom_polygon(data = map.df.merge, aes(x = long, y = lat, group = group,
        fill = coeff), color = NA) + 
  theme_minimal(base_size = 8) + coord_map() +
  scale_fill_distiller(palette = "YlOrRd") +
  ggtitle("Fixed Effect Coefficients for X4313314_Norm on 
          All Cause Hospitalization for Kids - Within County")

# Trying Google Maps Overlay
map <- get_map("WA State", zoom = 7, maptype = "roadmap")
p <- ggmap(map)
p + geom_polygon(data = map.df.merge, aes(x = long, y = lat, group = group,
  fill = coeff), color = "grey", size=0.1, alpha=0.6) +
  scale_fill_distiller(palette = "YlOrRd") + 
  ggtitle("WA State Counties-All Cause Hospitalization") +
  theme_minimal(base_size = 8)

```


### Now Running Fixed Effects for Specific Type of Diagnosis

#### Single-Level CCS - Diagnoses: [Codes](https://www.hcup-us.ahrq.gov/toolssoftware/ccs/CCSCategoryNames_FullLabels.pdf)

#### Diagnoses related to Respiratory Diseases

122 Pneumonia (except that caused by tuberculosis or sexually transmitted disease)
123 Influenza
124 Acute and chronic tonsillitis
125 Acute bronchitis
126 Other upper respiratory infections
127 Chronic obstructive pulmonary disease and bronchiectasis
128 Asthma
129 Aspiration pneumonitis; food/vomitus
130 Pleurisy; pneumothorax; pulmonary collapse
131 Respiratory failure; insufficiency; arrest (adult)
132 Lung disease due to external agents
133 Other lower respiratory disease
134 Other upper respiratory disease

```{r, message=FALSE, warning=FALSE}
# Filtering only Diagnoses related to Respiratory Diseases
hosp.resp <- hosp.filtered %>%
  select(ayear, amonth, zip, pstco2, AgeGroup, urban_rural, hospitalizations, dxccs1) %>%
  filter(dxccs1 >= 122 & dxccs1 <= 134)

# Aggregating
hosp.resp <- hosp.resp[c("ayear","amonth","pstco2","hospitalizations","AgeGroup")]
hosp.resp <- aggregate(hospitalizations~ayear+amonth+pstco2+AgeGroup, data=hosp.resp, sum)

# Joining population for normalization
hosp.resp <- merge(hosp.resp, census.clean, by.x=c("ayear","amonth","pstco2","AgeGroup"),
                       by.y=c("year","month","FIPS","AgeGroup"))
hosp.resp <- mutate(hosp.resp, resp_hosp_norm = hospitalizations/Population)
names(hosp.resp)[1] <- 'year'
names(hosp.resp)[2] <- 'month'
names(hosp.resp)[3] <- 'FIPS'

# Respiratory Hospitalizations
resp.merge <- merge(hosp.resp, vaccine.filtered, by=c("year","month","FIPS"))
resp.merge$t_index <- interaction(resp.merge$year, resp.merge$month,sep=".")
```

```{r, message=FALSE, warning=FALSE}
# Plotting the data
ggplot(resp.merge , aes(X4313314_Norm,resp_hosp_norm)) + 
  geom_point(alpha=0.1) + geom_smooth() + theme_minimal(base_size = 8) +
  ggtitle("Hospitalizations related to Respiratory Problems")

xyplot(resp_hosp_norm ~ X4313314_Norm | factor(County), data=resp.merge, 
       pch=19, type=c("p","r"), alpha = 0.1, 
       main=list(label="Hospitalizations related to Respiratory Problems By County",cex=0.8),
       sub=list(label="11",cex=0.1),
       par.settings=simpleTheme(col="#59ABE3", col.line="red"),
       par.strip.text=list(cex=0.5))
```

```{r, message=FALSE, warning=FALSE}
# Fixed Effects Respiratory related Hospitalization
fe.resp <- plm(resp_hosp_norm ~ X4313314_Norm, data = resp.merge, 
               index = c("FIPS", "t_index"), model = "within", effect = "individual")

summary(fe.resp)
summary(fixef(fe.resp))
fixef.resp <- matrix(fixef(fe.resp))
fixef.resp.df <- as.data.frame(cbind(sort(unique(resp.merge$FIPS)), fixef.resp))
colnames(fixef.resp.df) <- c("FIPS","coeff")
fixef.resp.df$FIPS <- as.character(fixef.resp.df$FIPS)
fixef.resp.df <- mutate(fixef.resp.df, id=substr(FIPS, 3,5))
fixef.resp.df <- fixef.resp.df[c("coeff",'id')]

```

```{r, message=FALSE, warning=FALSE}
# Choropleth for Respiratory related diagnoses
map.df.merge <- merge(map.df,fixef.resp.df, by="id") 
ggplot() +
  geom_polygon(data = map.df.merge, aes(x = long, y = lat, group = group,
        fill = coeff), color = NA) + 
  theme_minimal(base_size = 8) + coord_map() +
  scale_fill_distiller(palette = "YlOrRd") +
  ggtitle("Fixed Effect Coefficients for X4313314_Norm on 
          Respiratory related Hospitalization for Kids - Within County")

```


#### Diagnoses not related to Vaccinations 

226 Fracture of neck of femur (hip)
227 Spinal cord injury
228 Skull and face fractures
229 Fracture of upper limb
230 Fracture of lower limb
231 Other fractures
232 Sprains and strains
233 Intracranial injury
234 Crushing injury or internal injury
235 Open wounds of head; neck; and trunk
236 Open wounds of extremities
237 Complication of device; implant or graft
238 Complications of surgical procedures or medical care
239 Superficial injury; contusion
240 Burns
241 Poisoning by psychotropic agents
242 Poisoning by other medications and drugs
243 Poisoning by nonmedicinal substances
244 Other injuries and conditions due to external causes

```{r, message=FALSE, warning=FALSE}
# Filtering only Diagnoses related to Non-Respiratory Generic Diseases
hosp.general <- hosp.filtered %>%
  select(ayear, amonth, zip, pstco2, AgeGroup, urban_rural, hospitalizations, dxccs1) %>%
  filter(dxccs1 >= 226 & dxccs1 <= 244)

# Aggregating
hosp.general <- hosp.general[c("ayear","amonth","pstco2","hospitalizations","AgeGroup")]
hosp.general <- aggregate(hospitalizations~ayear+amonth+pstco2+AgeGroup, 
                          data=hosp.general, sum)

# Joining population for normalization
hosp.general <- merge(hosp.general, census.clean, 
                      by.x=c("ayear","amonth","pstco2","AgeGroup"),
                      by.y=c("year","month","FIPS","AgeGroup"))
hosp.general <- mutate(hosp.general, gen_hosp_norm = hospitalizations/Population)
names(hosp.general)[1] <- 'year'
names(hosp.general)[2] <- 'month'
names(hosp.general)[3] <- 'FIPS'

# Respiratory Hospitalizations
gen.merge <- merge(hosp.general, vaccine.filtered, by=c("year","month","FIPS"))
gen.merge$t_index <- interaction(gen.merge$year, gen.merge$month,sep=".")
```

```{r, message=FALSE, warning=FALSE}
# Plotting the data
ggplot(gen.merge , aes(X4313314_Norm,gen_hosp_norm)) + 
  geom_point(alpha=0.1) + geom_smooth() + theme_minimal(base_size = 8) +
  ggtitle("Generic Hospitalizations")

xyplot(gen_hosp_norm ~ X4313314_Norm | factor(County), data=gen.merge, 
       pch=19, type=c("p","r"), alpha = 0.1, 
       main=list(label="Generic Hospitalizations By County",cex=0.8),
       sub=list(label="11",cex=0.1),
       par.settings=simpleTheme(col="#59ABE3", col.line="red"),
       par.strip.text=list(cex=0.5))
```

```{r, message=FALSE, warning=FALSE}
# Fixed Effects - Generic Hospitalizations
fe.gen <- plm(gen_hosp_norm ~ X4313314_Norm, data = gen.merge, 
              index = c("FIPS", "t_index"), model = "within", effect = "individual")
summary(fe.gen)
summary(fixef(fe.gen))
fixef.gen <- matrix(fixef(fe.gen))
fixef.gen.df <- as.data.frame(cbind(sort(unique(gen.merge$FIPS)), fixef.gen))
colnames(fixef.gen.df) <- c("FIPS","coeff")
fixef.gen.df$FIPS <- as.character(fixef.gen.df$FIPS)
fixef.gen.df <- mutate(fixef.gen.df, id=substr(FIPS,3,5))
fixef.gen.df <- fixef.gen.df[c("coeff",'id')]

```

```{r, message=FALSE, warning=FALSE}
# Choropleth for Respiratory related diagnoses
map.df.merge <- merge(map.df,fixef.gen.df, by="id") 
ggplot() +
  geom_polygon(data = map.df.merge, aes(x = long, y = lat, group = group,
        fill = coeff), color = NA) + 
  theme_minimal(base_size = 8) + coord_map() +
  scale_fill_distiller(palette = "YlOrRd") +
  ggtitle("Fixed Effect Coefficients for X4313314_Norm on 
          Non-Respiratory related Generic Hospitalization for Kids - Within County")

```
