---
title: "HCUP - Models"
author: "Akashdeep Jaswal"
date: "June 8, 2016"
output: word_document
---

```{r, results='hide',message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results='hide',message=FALSE, warning=FALSE}

#Installing required packages
install.packages("plm",repos = "http://cran.us.r-project.org") 
install.packages("data.table",repos = "http://cran.us.r-project.org")

#Loading packages
library(plm) #For Running Fixed Effects Models

```

```{r}
# Reading the Cleaned data
all.data <- read.csv('/Users/ajaswal/Desktop/AkashPersonal/hcup-population/Final_Data_group1.csv')
all.data <- data.frame(all.data)

all.data <- subset(all.data, select = -c(X))

# Creating an interaction between Year and Month for Time Index for the random effects modeling
all.data <- transform(all.data, t_index=interaction(year,month,sep="."))

all.data <- all.data[order(all.data$County,all.data$t_index),c("County","t_index", "AgeGroup", "Population", "Patients_Count", "Patients_norm", "All_Vaccinated", "X4313314_SERIES", "X4313314_SERIES_Norm")]
head(all.data)

```

## Including Plots

You can also embed plots, for example:

```{r}
# Model1: Random Effects (Error Components) w/ Individual Effect (County) 

random.effects <- plm(Patients_norm ~ X4313314_SERIES_Norm, data = all.data, index = c("County", "t_index"), model = "random", effect="individual")
print(random.effects)
summary(random.effects)
# plot(random.effects)
```

Interpretation of Model 1: The coefficient for the X4313314_SERIES term is negative (-0.00378155) so that indicates that for every 1% increase in the series vaccination completion for all the vaccinated kids below the age of 4 years, the all cause hospitalization rate normalized by the kids population of the county goes down by 0.003. The results are statistically significant hence showing the expected relationship. 


```{r}
# Model 2: Fixed Effects w/ "Within" Model

# In the previous model, there was just one intercept but we got an error term for the individual. 
# To create a model with different intercepts for every individual county we build a fixed effects model 

fixed.effects <- plm(Patients_norm ~ X4313314_SERIES_Norm, data = all.data, index = c("County", "t_index"), model = "within", effect = "individual")
summary(fixed.effects)
summary(fixef(fixed.effects))

```

Observations of Model 2: The adjusted R-squared value has slightly increased indicating a slightly better fit from the data. Let's try inclusion of interaction terms between % X4313314_SERIES_Norm Vaccinated and the individual County in Model 3. 

```{r}
fixed.effects2 <- plm(Patients_norm ~ X4313314_SERIES_Norm*factor(County), data = all.data, index = c("County", "t_index"), model = "within", effect = "individual")
summary(fixed.effects2)
summary(fixef(fixed.effects2))

```

Observations from Model 3: The Adjusted R-squared has improved to indicate a better fit. Most Interaction term coefficients are statistically significant hinting at the fact that every county has a different receptive affinity for vaccinations. 

```{r}

# Shapefile Obtained from Gov Census - https://www.census.gov/geo/maps-data/data/cbf/cbf_counties.html

# Setting up Mapping 

#install.packages('rgdal', repos = "http://cran.us.r-project.org")
library(rgdal)
# install.packages("scales")
library(scales)
library(ggplot2)
library(ggmap)
library(dplyr)
# install.packages("rgeos")
library(rgeos)
library(broom)
# install.packages("Cairo")
library(Cairo)
library(maptools)

# Reading the Spatial Data Frame
map.shapefile <- readOGR(dsn = ".", layer = "cb_2015_us_county_500k")
map.df <- fortify(map.shapefile, region="GEOID") 

map <- get_map("WA State", zoom = 7, maptype = "roadmap")
p <- ggmap(map)
p



```
